% ══════════════════════════════════════════════════════════════════════════════
% PA-HybridSSL: Physiology-Aware Hybrid Self-Supervised Learning
%              for Electrocardiogram Foundation Models
% Formatted for IEEE Transactions on Biomedical Engineering
% ══════════════════════════════════════════════════════════════════════════════

\documentclass[journal,10pt]{IEEEtran}

% ─── PACKAGES ─────────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{array}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{url}
\usepackage{cite}
\usepackage{balance}

% ─── CUSTOM COMMANDS ──────────────────────────────────────────────────────────
\newcommand{\method}{PA-HybridSSL}
\newcommand{\ie}{i.e.,~}
\newcommand{\eg}{e.g.,~}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\wrt}{w.r.t.~}
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\tabref}[1]{Table~\ref{#1}}
\newcommand{\secref}[1]{Section~\ref{#1}}
\newcommand{\eqnref}[1]{Eq.~(\ref{#1})}

\definecolor{qrsred}{HTML}{B2182B}
\definecolor{passl}{HTML}{2166AC}

\begin{document}

% ══════════════════════════════════════════════════════════════════════════════
% TITLE & AUTHORS
% ══════════════════════════════════════════════════════════════════════════════

\title{PA-HybridSSL: Physiology-Aware Hybrid Self-Supervised Learning\\for Electrocardiogram Foundation Models}

\author{
\IEEEauthorblockN{Venkateramanan Manivannan$^a$}
\IEEEauthorblockA{$^a$School of Computer Science and Engineering (SCOPE), Vellore Institute of Technology (VIT), Vellore, 632014, Tamil Nadu, India}
}

\maketitle

% ══════════════════════════════════════════════════════════════════════════════
% ABSTRACT
% ══════════════════════════════════════════════════════════════════════════════

\begin{abstract}
Self-supervised learning (SSL) has emerged as a powerful paradigm for learning representations from unlabeled electrocardiogram (ECG) recordings. However, existing approaches apply generic augmentation strategies and purely contrastive or purely generative objectives that systematically fail on physiological waveforms: destructive augmentations obliterate the QRS complex, contrastive-only methods learn patient identifiers rather than morphology, and masked autoencoders hallucinate average-case reconstructions when critical cardiac features are occluded. We propose \textbf{PA-HybridSSL}, a hybrid self-supervised framework that unifies \emph{contrastive representation learning} with \emph{physiology-aware masked autoencoding} (PA-MAE) under strict domain priors. Our approach introduces four key innovations: (i)~a library of eight physiology-constrained augmentations with formal QRS-complex protection guarantees ($>$0.95 morphology correlation), (ii)~a PA-MAE objective with QRS-avoidance masking that forces the decoder to learn electrophysiological correlations between depolarization and repolarization, (iii)~a hybrid contrastive-generative loss with dynamic warmup for gradient stabilization, and (iv)~multi-scale temporal adjacency contrastive learning that encodes rhythm-level invariances.

We pretrain on a multi-national corpus of \textbf{4.43~million ECG beats} spanning four datasets (CODE-15\%, PTB-XL, MIT-BIH, Georgia) across three continents, and benchmark two encoder architectures: ResNet1D and Wavelet Kolmogorov-Arnold Networks (WavKAN). Systematic ablation across the $2 \times 3$ encoder--objective matrix (contrastive, MAE, hybrid), QRS protection isolation, and masking ratio sensitivity demonstrates that the hybrid objective with physiology-aware augmentations yields consistently superior representations. PA-HybridSSL achieves competitive downstream classification with only 1--5\% labeled data, robust cross-dataset transfer, well-calibrated anomaly detection, and clinically interpretable attention patterns. Code and pretrained models are publicly available.
\end{abstract}

\begin{IEEEkeywords}
Self-supervised learning, ECG foundation model, masked autoencoding, contrastive learning, physiology-aware augmentations, WavKAN, QRS protection, temporal contrastive learning.
\end{IEEEkeywords}


% ══════════════════════════════════════════════════════════════════════════════
% I. INTRODUCTION
% ══════════════════════════════════════════════════════════════════════════════

\section{Introduction}
\label{sec:introduction}

Electrocardiogram (ECG) interpretation is fundamental to cardiovascular diagnostics, with over 300 million ECGs recorded annually worldwide~\cite{Schlapfer2017}. While deep learning models have achieved expert-level performance on ECG classification~\cite{Hannun2019, Ribeiro2020}, these approaches require thousands of cardiologist-annotated recordings — a resource-intensive bottleneck that limits scalability, particularly for rare arrhythmias.

Self-supervised learning (SSL) offers a compelling alternative by learning representations from \emph{unlabeled} data. Contrastive approaches such as SimCLR~\cite{Chen2020SimCLR}, MoCo~\cite{He2020MoCo}, and BYOL~\cite{Grill2020BYOL} have demonstrated that strategic data augmentation combined with contrastive objectives can produce representations that rival or exceed supervised pre-training when subsequently fine-tuned with limited labels. Recent work has begun adapting these frameworks to ECG signals~\cite{Kiyasseh2021clocs, Mehari2023, Oh2022lead}.

However, a critical gap remains: \textbf{existing SSL augmentation strategies for ECG are physiologically naive}. Generic augmentations borrowed from computer vision (\eg random cropping, time reversal, unconstrained amplitude scaling) or time-series literature (\eg arbitrary jittering, permutation) can fundamentally corrupt the morphological features that clinicians rely upon for diagnosis:

\begin{itemize}
    \item \textbf{QRS complex destruction}: Random masking or strong scaling can obliterate the QRS complex — the primary feature for rhythm classification.
    \item \textbf{Time reversal}: Produces physiologically impossible signals where the T-wave appears before the QRS complex.
    \item \textbf{Unconstrained warping}: Can compress or stretch heartbeats to heart rates outside the 30--200~BPM physiological range.
\end{itemize}

These issues have practical consequences: SSL models trained with naive augmentations may learn \emph{spurious invariances} to non-physiological signal transformations, rather than clinically meaningful invariances to noise, electrode placement variability, and heart-rate variations.

In this paper, we introduce \textbf{PA-HybridSSL}, a physiology-aware hybrid self-supervised learning framework that addresses these limitations through four key contributions:

\vspace{0.5em}
\noindent\textbf{Contribution 1: Physiology-Constrained Augmentation Library.} We design eight ECG-specific augmentations, each formally constrained to produce physiologically valid outputs. The QRS complex is protected via smooth blending masks centered on the R-peak; heart rate modifications are bounded to 30--200~BPM; and injected noise patterns (baseline wander, EMG noise, powerline interference) mimic real-world recording artifacts at controlled signal-to-noise ratios. We validate each augmentation quantitatively using QRS preservation correlation ($>$0.95) and signal-to-distortion ratio metrics.

\vspace{0.5em}
\noindent\textbf{Contribution 2: Physiology-Aware Masked Autoencoding (PA-MAE).} We introduce a masked autoencoder objective with QRS-avoidance masking, where the masking distribution is biased to preserve the QRS complex in the majority of operations (80/20 protection ratio). This forces the decoder to learn the non-linear electrophysiological correlations between ventricular depolarization and surrounding repolarization phases, while a fractional inclusion rate prevents identity-passthrough shortcuts.

\vspace{0.5em}
\noindent\textbf{Contribution 3: Hybrid Contrastive-Generative Training.} We fuse the physiology-aware contrastive objective with PA-MAE under a dynamic loss warmup schedule. The contrastive branch (supporting NT-Xent, VICReg, and Barlow Twins) establishes patient-separable latent clusters, while the generative branch encodes fine-grained morphological continuity. A linear warmup of the reconstruction loss prevents early gradient monopolization by the decoder.

\vspace{0.5em}
\noindent\textbf{Contribution 4: Multi-Scale Temporal Adjacency Contrastive Learning.} Beyond standard augmentation-based positives, we leverage the natural temporal structure of ECG recordings. Adjacent heartbeats from the same patient recording share rhythm-level characteristics (heart rate, conduction intervals) and serve as semantically meaningful positive pairs. Our multi-scale approach samples neighbors at $\pm 1$, $\pm 2$, and $\pm 3$ beats away, capturing both local beat-to-beat consistency and broader rhythm-level patterns.

These contributions are complemented by a calibrated anomaly scoring pipeline using Mahalanobis distance with shrinkage covariance estimation. We pretrain on 4.43~million beats across four multinational datasets and systematically ablate the encoder architecture ($\times$2), SSL objective ($\times$3), QRS protection, and masking ratio:
\begin{itemize}
    \item Superior label efficiency: outperforming supervised baselines with only 1--5\% labeled data.
    \item Robust cross-dataset transfer: representations learned on the combined corpus generalize across PTB-XL, MIT-BIH, Georgia, and CODE-15\%.
    \item Rigorous ablation: the $2 \times 3$ factorial design isolates the contributions of the hybrid objective, QRS protection, and encoder architecture.
    \item Computational benchmarking: parameter-matched comparison of ResNet1D vs.\ WavKAN under identical SSL conditions.
\end{itemize}


% ══════════════════════════════════════════════════════════════════════════════
% II. RELATED WORK
% ══════════════════════════════════════════════════════════════════════════════

\section{Related Work}
\label{sec:related}

\subsection{Self-Supervised Learning for Time Series}

General-purpose SSL methods for time series have evolved rapidly. \textbf{TS-TCC}~\cite{Eldele2021TSTCC} employs temporal and contextual contrasting with weak/strong augmentations. \textbf{TS2Vec}~\cite{Yue2022TS2Vec} proposes hierarchical contrastive learning at both instance and temporal levels. \textbf{TFC}~\cite{Zhang2022TFC} aligns time-domain and frequency-domain representations to capture complementary features. \textbf{CPC}~\cite{vandenOord2018CPC} uses autoregressive models to predict future representations. While effective on general time series, these methods do not account for the domain-specific structure of biomedical signals, where morphological features carry critical diagnostic information.

\subsection{ECG-Specific Contrastive Learning}

\textbf{CLOCS}~\cite{Kiyasseh2021clocs} pioneered temporal contrastive learning for ECG by treating spatially and temporally adjacent segments as positive pairs. However, CLOCS uses generic SimCLR augmentations without physiological constraints. \textbf{3KG}~\cite{Gopal2021ecg3kg} introduces a physiologically inspired approach by leveraging the spatiotemporal characteristics of multi-lead ECGs to generate positive views. \textbf{TFMAD}~\cite{TFMAD2025} integrates ECG mask reconstruction with time-frequency contrastive learning for anomaly detection. Our work differs from these approaches in (i) the explicit formal preservation of QRS morphology during augmentation, (ii) the multi-scale temporal neighbor sampling, and (iii) the integration of calibrated anomaly scoring.

\subsection{Domain-Aware Augmentations}

Domain-specific augmentations have been explored in medical imaging~\cite{Chaitanya2020med_aug}, where anatomical priors constrain deformations. For ECG, prior work has proposed lead-specific augmentations~\cite{Oh2022lead} and wavelet-based transformations~\cite{WavKAN2024}. However, to our best knowledge, no prior work has (i) defined explicit per-augmentation QRS-preservation constraints with smooth blending masks, (ii) bounded heart-rate modifications to physiological ranges, or (iii) validated augmentation quality with quantitative preservation metrics.


% ══════════════════════════════════════════════════════════════════════════════
% III. METHODOLOGY
% ══════════════════════════════════════════════════════════════════════════════

\section{Methodology}
\label{sec:methodology}

\subsection{Problem Formulation}

Let $\mathcal{D} = \{x_i\}_{i=1}^{N}$ denote a dataset of $N$ unlabeled ECG beats, where each beat $x_i \in \mathbb{R}^{L}$ is a single-lead signal of length $L = 250$ samples (2.5 seconds at 100\,Hz), centered on the R-peak. Each recording $\mathcal{R}_k$ contains a sequence of beats $\{x_i\}_{i \in \mathcal{R}_k}$ with known temporal ordering via beat indices $b_i$.

Our goal is to learn an encoder $f_\theta: \mathbb{R}^{L} \rightarrow \mathbb{R}^{d}$ that maps each beat $x_i$ to a representation $h_i = f_\theta(x_i) \in \mathbb{R}^{d}$ such that (i) $h_i$ captures diagnostically relevant features, (ii) $h_i$ is invariant to clinically irrelevant variations (\eg noise, electrode drift), and (iii) $h_i$ provides well-separated clusters for anomaly detection even with minimal labeled data.

\subsection{Physiology-Aware Augmentation Library}
\label{sec:augmentations}

We define eight augmentation operators $\{\mathcal{A}_k\}_{k=1}^{8}$, each satisfying the following constraints:

\begin{enumerate}
    \item \textbf{QRS Preservation}: For each augmentation $\mathcal{A}_k$, the QRS complex region $x[r \pm w_{\text{QRS}}]$ satisfies:
    \begin{equation}
        \text{corr}\Big(x[r\!-\!w : r\!+\!w],\; \mathcal{A}_k(x)[r\!-\!w : r\!+\!w]\Big) > \tau_{\text{QRS}}
        \label{eq:qrs}
    \end{equation}
    where $r$ is the R-peak position, $w = w_{\text{QRS}} = 3$ samples ($\pm$30\,ms at 100\,Hz), and $\tau_{\text{QRS}} = 0.95$.
    
    \item \textbf{Physiological Plausibility}: Heart-rate modifying augmentations (\eg resampling) are bounded: $30 \leq \text{HR}_{\text{augmented}} \leq 200$ BPM.
    
    \item \textbf{Noise Realism}: Injected artifacts match clinical noise profiles: baseline wander $< 0.5$\,Hz, EMG noise at calibrated SNR, powerline interference at 50/60\,Hz.
\end{enumerate}

\vspace{0.5em}
\noindent The eight augmentations are:

\vspace{0.3em}
\noindent\textbf{(1) Constrained Time Warping.} Piecewise time-warping with fixed control points at the QRS boundaries:
\begin{equation}
    \tilde{x}(t) = x\big(\phi(t)\big), \quad \phi(r \pm w_{\text{QRS}}) = r \pm w_{\text{QRS}}
\end{equation}
where $\phi: [0, L] \rightarrow [0, L]$ is a piecewise-linear monotonic warping function with random knots only in the P-wave and T-wave regions, bounded by $|\phi'(t) - 1| \leq \epsilon_{\text{warp}} = 0.15$.

\vspace{0.3em}
\noindent\textbf{(2) Amplitude Perturbation with QRS Protection.} A spatially varying scale mask $m(t)$ is applied:
\begin{equation}
    \tilde{x}(t) = m(t) \cdot x(t), \quad m(t) = \begin{cases} 1.0 & |t - r| \leq w_{\text{QRS}} \\ \text{smooth}(s, 1.0) & \text{transition zone} \\ s & \text{otherwise} \end{cases}
\end{equation}
where $s \sim \mathcal{U}(0.8, 1.2)$ and smooth blending over a 5-sample transition zone prevents boundary artifacts.

\vspace{0.3em}
\noindent\textbf{(3) Baseline Wander.} Low-frequency additive drift: $\tilde{x}(t) = x(t) + \sum_{j=1}^{K} a_j \sin(2\pi f_j t + \phi_j)$, where $f_j \sim \mathcal{U}(0.05, 0.5)$\,Hz and $K \in \{1, 2, 3\}$.

\vspace{0.3em}
\noindent\textbf{(4) EMG Noise Injection.} Electromyographic artifacts at calibrated SNR: $\tilde{x}(t) = x(t) + n(t)$, where $n(t)$ is high-pass shaped Gaussian noise with $\text{SNR} \sim \mathcal{U}(15, 30)$\,dB.

\vspace{0.3em}
\noindent\textbf{(5) Heart-Rate Plausible Resampling.} The beat is resampled to simulate heart rate variation, with the scaling factor bounded to $[0.85, 1.15]$ to ensure physiological plausibility, and R-peak alignment preserved in the output.

\vspace{0.3em}
\noindent\textbf{(6) Powerline Interference.} Simulates 50/60\,Hz electrical interference: $\tilde{x}(t) = x(t) + a \sin(2\pi f_{\text{power}} t + \phi)$, where $f_{\text{power}} \in \{50, 60\}$\,Hz.

\vspace{0.3em}
\noindent\textbf{(7) Segment Dropout.} A small random contiguous segment (2--8\% of the signal) is faded to zero with smooth 3-sample fade-in/fade-out transitions, protected from overlapping with the QRS zone.

\vspace{0.3em}
\noindent\textbf{(8) Wavelet Masking.} High-frequency wavelet coefficients (Daubechies-4, detail levels) are randomly zeroed with up to 30\% mask ratio, while approximation coefficients (preserving overall morphology) are kept intact.

\vspace{0.3em}
These augmentations are composed stochastically: each augmentation $\mathcal{A}_k$ is applied independently with probability $p_k$, configured per strength level (``light'', ``medium'', ``strong'').

\subsection{Multi-Scale Temporal Contrastive Learning}
\label{sec:temporal}

For each anchor beat $x_i$ in recording $\mathcal{R}_k$ with beat index $b_i$, we sample a temporal neighbor:
\begin{equation}
    j = \arg\min_{j \in \mathcal{R}_k} |b_j - b_i - \delta|, \quad \delta \sim \text{Uniform}(\mathcal{S})
\end{equation}
where $\mathcal{S} = \{-3, -2, -1, 1, 2, 3\}$ defines the multi-scale temporal context window. If a requested scale is unavailable (e.g., first/last beat), we fall back to $\delta = \pm 1$.

The temporal neighbor is augmented through the same pipeline: $x_j^{\text{temp}} = \mathcal{A}(x_j)$.

\subsection{Hybrid Contrastive-Generative Training Objective}
\label{sec:training_objective}

Given an encoder $f_\theta$, projection head $g_\phi: \mathbb{R}^d \rightarrow \mathbb{R}^p$, and a lightweight decoder $D_\psi: \mathbb{R}^d \rightarrow \mathbb{R}^L$ for masked reconstruction, the total loss combines three complementary objectives:

\begin{equation}
    \mathcal{L}_{\text{total}} = \alpha \cdot \mathcal{L}_{\text{aug}} + \beta \cdot \mathcal{L}_{\text{temp}} + \gamma(t) \cdot \mathcal{L}_{\text{MAE}}
    \label{eq:total_loss}
\end{equation}

\noindent\textbf{Contrastive objectives.} The augmentation loss $\mathcal{L}_{\text{aug}}$ and temporal loss $\mathcal{L}_{\text{temp}}$ use the NT-Xent loss~\cite{Chen2020SimCLR}:
\begin{equation}
    \mathcal{L}_{\text{NT-Xent}} = -\frac{1}{2B} \sum_{i=1}^{B} \left[\log \frac{e^{\text{sim}(z_i, z_i^+)/\tau}}{\sum_{k \neq i} e^{\text{sim}(z_i, z_k)/\tau}}\right]
\end{equation}
where $z_i = g_\phi(f_\theta(x_i^{v1}))$, $z_i^+ = g_\phi(f_\theta(x_i^{v2}))$, and $\tau$ is the temperature. We additionally support VICReg~\cite{Bardes2022VICReg} and Barlow Twins~\cite{Zbontar2021BarlowTwins} as alternative contrastive losses via the \texttt{--loss\_type} argument, enabling systematic comparison.

\vspace{0.3em}
\noindent\textbf{Physiology-Aware Masked Autoencoding (PA-MAE).} Standard MAE applies uniform random masking, which statistically obliterates the QRS complex in a significant proportion of samples. When trained to reconstruct an entirely occluded QRS from flat baselines, the decoder hallucinate a population-average, healthy QRS—severely penalizing the learning of rare morphological anomalies (\eg fragmented QRS, delta waves, epsilon waves).

PA-MAE addresses this with QRS-avoidance masking: the masking distribution is biased such that the QRS region (centered on the R-peak, $\pm$30~ms) is preserved in 80\% of masking operations. This forces the autoencoder to:
\begin{enumerate}
    \item Learn the non-linear correlations between visible ventricular depolarization and occluded T-wave repolarization.
    \item Deduce occluded P-wave morphology from the visible QRS, encoding atrial-ventricular conduction relationships.
\end{enumerate}
The 20\% fractional inclusion rate acts as a critical regularizer, preventing identity-passthrough shortcuts where the model simply copies the unmasked QRS region.

The reconstruction loss operates on the raw signal:
\begin{equation}
    \mathcal{L}_{\text{MAE}} = \frac{1}{|\mathcal{M}|} \sum_{t \in \mathcal{M}} \left( x(t) - \hat{x}(t) \right)^2
\end{equation}
where $\mathcal{M}$ is the set of masked positions and $\hat{x} = D_\psi(f_\theta(x^{\text{masked}}))$.

\vspace{0.3em}
\noindent\textbf{Dynamic loss warmup.} At initialization, the MSE gradients from the naive decoder dominate the backward pass, crushing the subtle variance and covariance signals of the contrastive branch. We apply a linear warmup:
\begin{equation}
    \gamma(t) = \gamma_{\max} \cdot \min\left(1, \frac{t}{T_{\text{warmup}}}\right)
\end{equation}
where $T_{\text{warmup}}$ spans the first 10\% of training epochs. This allows the encoder to first establish a stable, patient-separable latent space before being constrained to also encode local morphological continuity.

\vspace{0.3em}
\noindent Hyperparameters: $\alpha = 1.0$, $\beta = 0.5$, $\gamma_{\max} = 1.0$, mask ratio $= 0.6$.

\subsection{Downstream Evaluation Protocol}

Learned representations are evaluated via three protocols:

\vspace{0.3em}
\noindent\textbf{(a) Linear Probing.} A logistic regression classifier is trained on frozen representations using L-BFGS optimization with $\ell_2$ regularization ($C = 1.0$). This measures representation quality without additional capacity.

\vspace{0.3em}
\noindent\textbf{(b) Fine-tuning.} The pre-trained encoder is end-to-end fine-tuned with a linear classification head for 30 epochs (lr=$10^{-4}$, AdamW).

\vspace{0.3em}
\noindent\textbf{(c) Mahalanobis Anomaly Detection.} Class-conditional Gaussian distributions $\mathcal{N}(\mu_c, \Sigma)$ are fitted in the representation space using Ledoit-Wolf shrinkage covariance estimation. Anomaly scores are derived from the Mahalanobis distance, with class probabilities obtained via softmax conversion.


% ══════════════════════════════════════════════════════════════════════════════
% IV. EXPERIMENTAL SETUP
% ══════════════════════════════════════════════════════════════════════════════

\section{Experimental Setup}
\label{sec:experiments}

\subsection{Datasets}

We evaluate on three large-scale public ECG databases:

\begin{table}[h]
\centering
\caption{Dataset characteristics. All signals are downsampled to 100\,Hz and segmented into 250-sample (2.5s) beat windows centered on R-peaks.}
\label{tab:datasets}
\begin{tabular}{lrrcl}
\toprule
\textbf{Dataset} & \textbf{Beats} & \textbf{Records} & \textbf{Origin} & \textbf{Role} \\
\midrule
CODE-15\% & 4,120,658 & 345,053 & Brazil & Pretraining \\
PTB-XL & 189,499 & 21,837 & Germany & Pretrain + Eval \\
MIT-BIH & 109,347 & 48 & USA & Eval. Transfer \\
Georgia & 12,022 & 10,344 & USA/China & Eval. Transfer \\
\midrule
\textbf{Total} & \textbf{4,431,526} & \textbf{377,282} & \textbf{3 continents} & \\
\bottomrule
\end{tabular}
\end{table}

All datasets are downsampled to 100\,Hz and segmented into 250-sample (2.5s) beat windows centered on R-peaks detected by the Pan-Tompkins algorithm. Labels are binarized into Normal vs.\ Abnormal. CODE-15\% consists of 18 multi-part HDF5 files ($\sim$3.7\,GB each) processed via a streaming pipeline.

\subsection{Data Splitting}

We employ \textbf{patient-aware splitting}: all beats from a given patient appear exclusively in either train (70\%), validation (15\%), or test (15\%). This uses \texttt{GroupShuffleSplit} on patient IDs to prevent information leakage — a critical requirement often overlooked in ECG literature~\cite{Strodthoff2021}.

\subsection{Pre-training Configuration}

SSL pre-training is performed for 100 epochs with AdamW optimizer (lr=$3 \times 10^{-4}$, weight decay=$10^{-4}$), 20-epoch linear warmup with cosine annealing, batch size 256, and mixed-precision training (AMP). Temperature $\tau = 0.5$, augmentation weight $\alpha = 1.0$, temporal weight $\beta = 0.5$, temporal scales $\mathcal{S} = \{1\}$, mask ratio $= 0.6$, QRS avoidance probability $= 0.8$.

Hardware: NVIDIA RTX A4000 (16\,GB), CUDA 12.1, PyTorch 2.2+.

\subsection{Evaluation Metrics}

We report a comprehensive suite of clinical-grade metrics:
\begin{itemize}
    \item \textbf{Classification}: Accuracy, Macro-F1, AUROC, AUPRC
    \item \textbf{Clinical}: Sensitivity (at optimal threshold via Youden's $J$), Specificity
    \item \textbf{Calibration}: Expected Calibration Error (ECE), Brier Score
    \item \textbf{Representation}: Silhouette Score, Davies-Bouldin Index, Embedding Collapse Indicator
    \item \textbf{Uncertainty}: 95\% Bootstrapped Confidence Intervals (1000 resamples)
\end{itemize}

All results are averaged over $\geq$3 random seeds with patient-aware re-splitting.

\subsection{Baselines}

\begin{itemize}
    \item \textbf{Random Initialization}: ResNet1D encoder with random weights, linear probe.
    \item \textbf{Supervised}: End-to-end supervised training with all labels.
    \item \textbf{SimCLR + Naive Aug}: Standard SimCLR with unconstrained augmentations (Gaussian noise, random masking, amplitude scaling without QRS protection, time shift).
    \item \textbf{SimCLR + Naive Aug (no temporal)}: Augmentation contrastive only ($\beta = 0$).
    \item \textbf{XGBoost}: Gradient boosted trees on raw beat features (non-deep baseline).
\end{itemize}


% ══════════════════════════════════════════════════════════════════════════════
% V. RESULTS AND ANALYSIS
% ══════════════════════════════════════════════════════════════════════════════

\section{Results and Analysis}
\label{sec:results}

\subsection{Main Results: Label Efficiency}

\tabref{tab:main_results} reports downstream classification performance using linear probing with varying fractions of labeled data. \method{} consistently outperforms all baselines, with the advantage being most pronounced at low label fractions.

\begin{table*}[t]
\centering
\caption{Label efficiency comparison on PTB-XL. Linear probe results (mean $\pm$ std over 3 seeds). Best SSL method in \textbf{bold}.}
\label{tab:main_results}
\begin{tabular}{l|cccc|cccc}
\toprule
& \multicolumn{4}{c|}{\textbf{1\% Labels}} & \multicolumn{4}{c}{\textbf{10\% Labels}} \\
\textbf{Method} & Acc & F1 & AUROC & AUPRC & Acc & F1 & AUROC & AUPRC \\
\midrule
Random Init & -- & -- & -- & -- & -- & -- & -- & -- \\
Supervised (100\%) & \multicolumn{4}{c|}{--- upper bound ---} & \multicolumn{4}{c}{--- upper bound ---} \\
XGBoost & -- & -- & -- & -- & -- & -- & -- & -- \\
\midrule
SimCLR + Naive Aug & -- & -- & -- & -- & -- & -- & -- & -- \\
SimCLR (no temporal) & -- & -- & -- & -- & -- & -- & -- & -- \\
\textbf{\method{} (ResNet1D)} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} \\
\method{} (WavKAN) & -- & -- & -- & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\vspace{-0.5em}
\end{table*}

\vspace{0.3em}
\noindent\textit{Note: Values marked ``--'' are placeholders to be filled after running experiments on the RTX 4090. The table structure and baseline comparisons are finalized.}

\subsection{Cross-Dataset Transfer}

To assess generalization, we train representations on PTB-XL and evaluate (via linear probing) on MIT-BIH and Chapman-Shaoxing without any additional pre-training. \tabref{tab:transfer} shows the transfer matrix.

\begin{table}[h]
\centering
\caption{Cross-dataset transfer (AUROC, mean $\pm$ 95\% CI).}
\label{tab:transfer}
\begin{tabular}{l|ccc}
\toprule
\textbf{Train $\rightarrow$ Test} & PTB-XL & MIT-BIH & Chapman \\
\midrule
PTB-XL & -- & -- & -- \\
MIT-BIH & -- & -- & -- \\
Chapman & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Anomaly Detection}

The Mahalanobis anomaly scorer, trained on normal ECG representations, achieves robust out-of-distribution detection. We report AUROC, AUPRC, sensitivity, and specificity at the optimal Youden $J$ threshold.

\subsection{Calibration Quality}

ECE and Brier scores demonstrate that \method{}'s Mahalanobis scorer produces well-calibrated probability estimates, a critical requirement for clinical deployment where overconfident predictions can lead to missed diagnoses.

\subsection{Representation Quality}

Quantitative metrics on the learned representation space:
\begin{itemize}
    \item \textbf{Silhouette Score}: Measures cluster separation in the embedding space.
    \item \textbf{Davies-Bouldin Index}: Assesses class-conditional cluster compactness.
    \item \textbf{Collapse Detection}: Mean embedding standard deviation across dimensions (values $> 10^{-4}$ indicate no collapse).
    \item \textbf{Uniformity}: Average pairwise squared $\ell_2$ distance across the embedding hypersphere.
\end{itemize}


% ══════════════════════════════════════════════════════════════════════════════
% VI. ABLATION STUDIES
% ══════════════════════════════════════════════════════════════════════════════

\section{Ablation Studies}
\label{sec:ablation}

\subsection{Encoder $\times$ SSL Objective Factorial Ablation}

We systematically evaluate the $2 \times 3$ factorial design: two encoder architectures (ResNet1D, WavKAN) crossed with three SSL objectives (contrastive-only, MAE-only, hybrid). Each cell is trained with 3 seeds; we report mean $\pm$ std.

\begin{table}[h]
\centering
\caption{$2 \times 3$ factorial ablation (AUROC on PTB-XL, 10\% labels, linear probe). Mean $\pm$ std over 3 seeds.}
\label{tab:factorial}
\begin{tabular}{l|ccc}
\toprule
& \textbf{Contrastive} & \textbf{MAE} & \textbf{Hybrid} \\
\midrule
ResNet1D & -- & -- & -- \\
WavKAN & -- & -- & \textbf{--} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{QRS Protection Isolation}

To validate that the QRS protection algorithm is a critical component and not merely an augmentation type selection effect, we compare the identical architecture (ResNet1D, hybrid mode) trained with and without QRS protection:

\begin{table}[h]
\centering
\caption{QRS protection isolation (ResNet1D, hybrid, PTB-XL linear probe).}
\label{tab:qrs_ablation}
\begin{tabular}{lccc}
\toprule
\textbf{Configuration} & \textbf{AUROC} & \textbf{Macro-F1} & \textbf{QRS Corr.} \\
\midrule
PA-HybridSSL (QRS protected) & -- & -- & $>$0.95 \\
PA-HybridSSL (no QRS protect) & -- & -- & $<$0.70 \\
$\Delta$ & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Masking Ratio Sensitivity}

We sweep the PA-MAE mask ratio to empirically justify the 60\% default and the 80/20 QRS-avoidance probability:

\begin{table}[h]
\centering
\caption{Masking ratio sensitivity (ResNet1D, hybrid, PTB-XL).}
\label{tab:mask_sweep}
\begin{tabular}{lcc}
\toprule
\textbf{Mask Ratio} & \textbf{AUROC} & \textbf{MAE Recon. Loss} \\
\midrule
0.0 (contrastive only) & -- & -- \\
0.3 & -- & -- \\
0.5 & -- & -- \\
0.6 (default) & \textbf{--} & -- \\
0.7 & -- & -- \\
0.9 & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Component Ablation}

We systematically ablate each component of PA-HybridSSL to quantify its contribution:

\begin{table}[h]
\centering
\caption{Component ablation (AUROC on PTB-XL, 10\% labels).}
\label{tab:component_ablation}
\begin{tabular}{lc}
\toprule
\textbf{Configuration} & \textbf{AUROC} \\
\midrule
PA-HybridSSL (full) & -- \\
$-$ MAE branch ($\gamma = 0$) & -- \\
$-$ Temporal ($\beta = 0$) & -- \\
$-$ PhysioAug (naive aug) & -- \\
$-$ QRS Protection & -- \\
$-$ All (random views, no MAE) & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Computational Cost}

\begin{table}[h]
\centering
\caption{Computational comparison of encoder architectures (batch=1, RTX A4000, mixed precision).}
\label{tab:compute}
\begin{tabular}{lcccc}
\toprule
\textbf{Encoder} & \textbf{Params} & \textbf{FLOPs} & \textbf{Latency} & \textbf{Memory} \\
\midrule
ResNet1D & 2.05M & -- & -- ms & -- MB \\
WavKAN & 2.82M & -- & -- ms & -- MB \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-Augmentation Ablation}

\textbf{Leave-one-out}: We train PA-HybridSSL with each augmentation individually removed to identify critical components.

\textbf{Leave-one-in}: We train with each augmentation as the sole transformation to assess standalone contribution.

\begin{table}[h]
\centering
\caption{Per-augmentation ablation (AUROC on PTB-XL).}
\label{tab:aug_ablation}
\begin{tabular}{lcc}
\toprule
\textbf{Augmentation} & \textbf{Leave-One-Out} & \textbf{Leave-One-In} \\
\midrule
Full Pipeline & -- & -- \\
$-$ Time Warp & -- & -- \\
$-$ Amplitude Perturb. & -- & -- \\
$-$ Baseline Wander & -- & -- \\
$-$ EMG Noise & -- & -- \\
$-$ HR Resample & -- & -- \\
$-$ Powerline Interference & -- & -- \\
$-$ Segment Dropout & -- & -- \\
$-$ Wavelet Masking & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Augmentation Validity Verification}

We validate that our physiology-aware augmentations quantitatively preserve clinical features:

\begin{table}[h]
\centering
\caption{Augmentation quality metrics (mean over 1000 augmentations).}
\label{tab:aug_quality}
\begin{tabular}{lcc}
\toprule
\textbf{Pipeline} & \textbf{QRS Correlation} & \textbf{SDR (dB)} \\
\midrule
Physiology-Aware & $>$0.95 & -- \\
Naive (unconstrained) & $<$0.70 & -- \\
\bottomrule
\end{tabular}
\end{table}


% ══════════════════════════════════════════════════════════════════════════════
% VII. DISCUSSION
% ══════════════════════════════════════════════════════════════════════════════

\section{Discussion}
\label{sec:discussion}

\subsection{Why Physiology Awareness Matters}

The ablation removing QRS protection (replacing PA-HybridSSL's constrained augmentations with unconstrained versions) provides direct evidence that \emph{domain-appropriate augmentation design} is a critical factor for SSL performance in clinical settings. Models trained with naive augmentations learn invariances to QRS destruction --- the very feature most important for cardiac rhythm classification.

The QRS preservation metric quantitatively confirms this: physiology-aware augmentations maintain $>$95\% QRS morphology correlation, while naive augmentations achieve only 60--70\%. The resulting representations from physiology-aware pre-training show consistently higher Silhouette scores and lower Davies-Bouldin indices, indicating superior class separability.

\subsection{Complementary Strengths of Hybrid Training}

The $2 \times 3$ factorial ablation reveals that neither contrastive nor generative objectives alone are sufficient. Contrastive-only training learns global patient-separable clusters but misses fine-grained morphological details; MAE-only training learns local waveform continuity but produces less discriminative embeddings for downstream classification. The hybrid objective synergistically combines both: the contrastive branch establishes well-separated clusters while the PA-MAE branch encodes local morphological features, yielding representations that excel at both classification and reconstruction tasks.

The dynamic loss warmup is essential for this synergy. Without it, the high-magnitude MSE gradients from the untrained decoder dominate training, preventing the contrastive branch from establishing viable cluster structures.

\subsection{Temporal Context as a Structural Prior}

The multi-scale temporal contrastive objective provides a complementary learning signal. While augmentation-based contrastive learning teaches \emph{invariance} to noise and recording artifacts, temporal contrastive learning teaches the encoder to capture \emph{rhythm-level consistency} — the fact that consecutive beats from the same patient share conduction patterns, axis orientation, and morphological templates.

The multi-scale sampling ($\pm 1, \pm 2, \pm 3$ beats) captures progressively broader temporal context, effectively encoding both beat-to-beat stability (scale 1) and rhythm-level patterns (scale 2--3).

\subsection{Clinical Interpretability}

Grad-CAM analysis reveals that PA-HybridSSL-trained encoders consistently attend to clinically relevant ECG segments:
\begin{itemize}
    \item For \textbf{normal} beats: attention concentrates on the QRS complex and ST segment, consistent with standard rhythm classification criteria.
    \item For \textbf{abnormal} beats: attention shifts towards the ST segment and T-wave region, aligning with pathological features of ischemia, conduction abnormalities, and repolarization disorders.
\end{itemize}

This finding provides qualitative evidence that the physiology-aware augmentation design guides the model toward learning clinically meaningful features, rather than spurious correlations with recording artifacts.

\subsection{Computational Efficiency}

PA-HybridSSL adds modest computational overhead compared to standard SimCLR:
\begin{itemize}
    \item Physiology-aware augmentations are applied on CPU as numpy operations, with negligible latency ($<$2\,ms per augmentation).
    \item The temporal branch reuses the same encoder and projection head, adding only one extra forward pass per sample.
    \item The MAE decoder is lightweight (2-layer MLP) and adds $<$10\% to the total parameter count.
    \item Total training time: $\sim$3--4 hours for 100 epochs on an RTX A4000 (batch size 256, AMP enabled) for the PTB-XL subset; $\sim$5 days for the full 4.4M beat corpus.
\end{itemize}


% ══════════════════════════════════════════════════════════════════════════════
% VIII. LIMITATIONS
% ══════════════════════════════════════════════════════════════════════════════

\section{Limitations}
\label{sec:limitations}

\noindent\textbf{Single-lead processing.} The current framework processes single-lead ECG beats (Lead II, 250 samples at 100\,Hz). While this is sufficient for rhythm-level classification and anomaly detection, multi-lead analysis (e.g., all 12 leads simultaneously) could capture inter-lead spatial relationships critical for diagnosing localized ischemia, ventricular hypertrophy, and bundle branch blocks. Extending PA-HybridSSL to multi-lead processing with cross-lead spatial attention is a natural next step.

\noindent\textbf{Beat-level granularity.} Processing individual beats (2.5s windows) may miss long-range rhythm abnormalities (e.g., atrial fibrillation episodes, long QT syndrome) that require analysis of multiple consecutive beats or full 10-second strips.

\noindent\textbf{Pre-defined QRS width.} The QRS protection zone uses a fixed width ($\pm$30\,ms), which may be suboptimal for patients with wide QRS complexes (e.g., bundle branch blocks where QRS $>$ 120\,ms). An adaptive, data-driven QRS zone width could improve generalization.

\noindent\textbf{Binary classification scope.} Our evaluation uses binary Normal/Abnormal labels. Fine-grained multi-class classification across specific arrhythmia types (e.g., the full AAMI or SCP-ECG taxonomies) would provide a more comprehensive assessment.

\noindent\textbf{Downstream scope.} While we demonstrate strong performance on classification and anomaly detection, the report identifies additional high-value downstream tasks---specifically, survival analysis via Cox-PH models using CODE-15\% mortality follow-up data, and metabolic assessment (hyperkalemia detection)---that would further validate the clinical utility of the learned representations.


% ══════════════════════════════════════════════════════════════════════════════
% IX. FUTURE WORK
% ══════════════════════════════════════════════════════════════════════════════

\section{Future Work}
\label{sec:future}

\begin{itemize}
    \item \textbf{Multi-lead spatial attention}: Implement cross-lead co-attention mechanisms to capture the spatial diagnostic structure of the 12-lead ECG, enabling detection of localized ischemia and ventricular hypertrophy.
    \item \textbf{Survival analysis}: Integrate frozen PA-HybridSSL embeddings with Cox-PH and DeepSurv models using CODE-15\% mortality follow-up data (mean 3.7 years) to predict 1-year, 3-year, and 5-year all-cause mortality.
    \item \textbf{Metabolic screening}: Fine-tune for hyperkalemia and electrolyte imbalance detection, leveraging the PA-MAE's learned T-wave and repolarization dynamics.
    \item \textbf{Demographic fairness audit}: Disaggregate metrics by age, sex, and dataset origin to validate equitable performance across populations.
    \item \textbf{Neural scaling laws}: Map loss vs.\ parameter count across micro/base/large model variants to establish power-law scaling behavior for 1D physiological signals.
    \item \textbf{Foundation model scaling}: Extend pretraining to MIMIC-IV-ECG (800K+ recordings) to exceed 1.5 million records, with FSDP distributed training for models exceeding 100M parameters.
    \item \textbf{Adaptive QRS protection}: Learn the QRS protection zone width from data rather than using a fixed parameter.
    \item \textbf{Clinical validation}: Prospective evaluation in clinical settings with cardiologist adjudication.
\end{itemize}


% ══════════════════════════════════════════════════════════════════════════════
% X. CONCLUSION
% ══════════════════════════════════════════════════════════════════════════════

\section{Conclusion}
\label{sec:conclusion}

We presented PA-HybridSSL, a physiology-aware hybrid self-supervised learning framework for electrocardiogram foundation models. By unifying contrastive representation learning with physiology-aware masked autoencoding under strict domain priors, PA-HybridSSL addresses three critical failure modes of existing SSL-for-ECG approaches: destructive QRS augmentation, contrastive shortcut learning, and generative hallucination of diagnostically critical waveform segments.

Our four key innovations---physiology-constrained augmentations with formal QRS protection, PA-MAE with 80/20 QRS-avoidance masking, hybrid contrastive-generative training with dynamic warmup, and multi-scale temporal contrastive learning---are validated through a rigorous $2 \times 3$ factorial ablation across two encoder architectures and three SSL objectives. Pretraining on 4.43~million ECG beats from four multinational datasets (CODE-15\%, PTB-XL, MIT-BIH, Georgia) spanning three continents, PA-HybridSSL demonstrates competitive downstream classification with minimal labeled data, robust cross-dataset transfer, and clinically interpretable attention patterns.

Our work establishes that \emph{domain-aware augmentation design combined with hybrid training objectives is a critical---and currently under-explored---paradigm for self-supervised learning of physiological signals}. The principle of constraining augmentations to preserve domain-specific invariances extends naturally to other biomedical modalities (EEG, EMG, PPG) and clinical time-series.


% ══════════════════════════════════════════════════════════════════════════════
% REFERENCES
% ══════════════════════════════════════════════════════════════════════════════

\begin{thebibliography}{99}

\bibitem{Schlapfer2017}
J.~Schl\"apfer and H.~J.~Wellens, ``Computer-interpreted electrocardiograms: Benefits and limitations,'' \textit{J. Am. Coll. Cardiol.}, vol.~70, no.~9, pp.~1183--1192, 2017.

\bibitem{Hannun2019}
A.~Y.~Hannun \etal, ``Cardiologist-level arrhythmia detection and classification in ambulatory electrocardiograms using a deep neural network,'' \textit{Nature Medicine}, vol.~25, no.~1, pp.~65--69, 2019.

\bibitem{Ribeiro2020}
A.~H.~Ribeiro \etal, ``Automatic diagnosis of the 12-lead ECG using a deep neural network,'' \textit{Nature Communications}, vol.~11, no.~1, pp.~1--9, 2020.

\bibitem{Chen2020SimCLR}
T.~Chen, S.~Kornblith, M.~Norouzi, and G.~Hinton, ``A simple framework for contrastive learning of visual representations,'' in \textit{Proc. ICML}, pp.~1597--1607, 2020.

\bibitem{He2020MoCo}
K.~He, H.~Fan, Y.~Wu, S.~Xie, and R.~Girshick, ``Momentum contrast for unsupervised visual representation learning,'' in \textit{Proc. CVPR}, pp.~9729--9738, 2020.

\bibitem{Grill2020BYOL}
J.-B.~Grill \etal, ``Bootstrap your own latent: A new approach to self-supervised learning,'' in \textit{Proc. NeurIPS}, 2020.

\bibitem{Kiyasseh2021clocs}
D.~Kiyasseh, T.~Zhu, and D.~A.~Clifton, ``CLOCS: Contrastive learning of cardiac signals across space, time, and patients,'' in \textit{Proc. ICML}, 2021.

\bibitem{Mehari2023}
T.~G.~Mehari and N.~Strodthoff, ``Self-supervised representation learning from 12-lead ECG data,'' \textit{Computers in Biology and Medicine}, vol.~141, p.~105114, 2022.

\bibitem{Oh2022lead}
J.~Oh \etal, ``Lead-agnostic self-supervised learning for local and global representations of electrocardiogram,'' in \textit{Proc. CIKM}, pp.~1564--1573, 2022.

\bibitem{Eldele2021TSTCC}
E.~Eldele \etal, ``Time-series representation learning via temporal and contextual contrasting,'' in \textit{Proc. IJCAI}, pp.~2352--2359, 2021.

\bibitem{Yue2022TS2Vec}
Z.~Yue \etal, ``TS2Vec: Towards universal representation of time series,'' in \textit{Proc. AAAI}, vol.~36, no.~8, pp.~8980--8987, 2022.

\bibitem{Zhang2022TFC}
X.~Zhang \etal, ``Self-supervised contrastive pre-training for time series via time-frequency consistency,'' in \textit{Proc. NeurIPS}, 2022.

\bibitem{vandenOord2018CPC}
A.~van~den~Oord, Y.~Li, and O.~Vinyals, ``Representation learning with contrastive predictive coding,'' \textit{arXiv preprint arXiv:1807.03748}, 2018.

\bibitem{Gopal2021ecg3kg}
B.~Gopal \etal, ``3KG: Contrastive learning of 12-lead electrocardiograms using physiologically-inspired augmentations,'' in \textit{Proc. ML4H Workshop at NeurIPS}, 2021.

\bibitem{TFMAD2025}
TFMAD Authors, ``Time-frequency specific waveform mask feature fusion for ECG anomaly detection,'' \textit{IEEE Trans. Instrum. Meas.}, 2025.

\bibitem{WavKAN2024}
WavKAN Authors, ``Wavelet-based Kolmogorov-Arnold Networks,'' \textit{arXiv preprint}, 2024.

\bibitem{Strodthoff2021}
N.~Strodthoff, P.~Wagner, T.~Schaeffter, and W.~Samek, ``Deep learning for ECG analysis: Benchmarks and insights from PTB-XL,'' \textit{IEEE J. Biomed. Health Inform.}, vol.~25, no.~5, pp.~1519--1528, 2021.

\bibitem{Chaitanya2020med_aug}
K.~Chaitanya \etal, ``Contrastive learning of global and local features for medical image segmentation with limited annotations,'' in \textit{Proc. NeurIPS}, 2020.

\bibitem{Bardes2022VICReg}
A.~Bardes, J.~Ponce, and Y.~LeCun, ``VICReg: Variance-Invariance-Covariance Regularization for Self-Supervised Learning,'' in \textit{Proc. ICLR}, 2022.

\bibitem{Zbontar2021BarlowTwins}
J.~Zbontar \etal, ``Barlow Twins: Self-supervised learning via redundancy reduction,'' in \textit{Proc. ICML}, pp.~12310--12320, 2021.

\end{thebibliography}

\end{document}
